---
title: "Using clumpedr: basic data analysis"
author: "Ilja J. Kocken"
date: 2019-01-31
output:
  rmarkdown::html_vignette:
    self_contained: no
vignette: >
  %\VignetteIndexEntry{Using clumpedr: basic data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  autodep = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

Once you have `clumpedr` installed (see the README), you can first load the libraries:

# load the packages that we use
```{r}
  # library(tidyverse)  # a few of the below and many more
  library(glue)      # optional, if you want to glue strings together
  library(dplyr)     # for pipes, mutate, summarise, etc.
  library(ggplot2)   # for plots!

  library(isoreader) # read in raw data
  library(clumpedr)  # this package! process clumped isotope data
```

# get your data from raw instrument data into R

## load data from a remote

First locate where you have your raw data files stored. This example uses a
linux machine, accessing a windows samba server.

Of course you can also just copy your files and paste them in a folder you
desire.

```{r}
folderstr <- "/run/user/1000/gvfs/smb-share:server=geofile02.geo.uu.nl,share=geo-labs/@RawData"
# read all did files
dids <- iso_read_dual_inlet(glue("{folderstr}/253pluskiel/Raw Data/Kiel Raw Data"),
                            cache = TRUE,
                            discard_duplicates = FALSE,
                            parallel = TRUE)
```

It's nice to save a cache/backup as an R data structure file, which we can read
in much faster.

```{r}
iso_save(dids, "out/dids.di.rds")
```

## load from the cache

Once we have saved the r data storage (`.rds`) file, we can load it much faster
than the raw data.

```{r}
dids  <- iso_read_dual_inlet("out/dids.di.rds")
```

# process the data!

Clean up the metadata and append the mass spectrometer name. This function
parses integers, numbers, etc. so that all the types are correct.

```{r}
didinfo  <- clean_did_info(dids, masspec = "MOTU")
```

## run the processing step-by-step

Note that normally, it's faster and smarter not to save the output of every
step as a separate tibble, but in this case we do it so that we can easily
inspect the results along the way. See the end of the vignette for the single
pipe.

### filter measurements of interest

First we filter out the measurements we want, based on the Method name.

```{r}
filt <- dids %>%
  # we can subset to some files of interest (e.g., based on a regular expression)
  # iso_filter_files(grepl("[0-9]{6}_129|[0-9]{6}_130", file_id)) %>%
  # all clumped runs
  filter_clumped()
```

### extract raw data

Then we extract the raw data. This gives it in a format where each row
is one cycle of either `standard` or `sample` gas, with initensities
44--49 as columns.

```{r}
rawd <- filt %>%
  # get all the raw data, per cycle from the dids
  iso_get_raw_data()
```

### disable failed cycles

This disables any cycles that have a sudden drop in pressure/intensity. With
the `genplot = TRUE` option, we tell it to create a default ggplot to inspect
which cycles have been disabled.

```{r}
disc <- rawd %>%
  # get rid of bad cycles
  disable_cycles(fac = 2, genplot = TRUE)
```

### HINT: look at plots interactively

Note that it is very nice to look at this plot---or any of the future
ones---interactively using `ggplotly`:

```{r}
library(plotly)
ggplotly(dynamicTicks = TRUE)
```

### background correction

Do a very simple background correction.

```{r}
bgds <- disc %>%
  # do a very simple background correction
  correct_backgrounds(factor = 0.82)
```

### spread match

Spread and match intensities. First we re-order the data into a wide format,
where sample and reference gas intensities are listed next to each other as
separate columns using the gather-unite-spread approach.

Then we compare reference gas to sample gas. With the `method="normal"`, this
would calculate the average of first and second cycles for the reference gas.
We can also use a work-in-progress linear interpolation (`method="linterp"`) to
match the mass 44 intensity of the reference gas to that of the sample gas and
apply this same shift to all the other masses. At present, it performs more
poorly than the regular calculation though, probably due to cycle elimination.

For example, we convert from the below:

| `file_id`      | type         | cycle | v44.mV | v45.mV | v46.mV | v47.mV | v54.mV |
| `"file_1.did"` | `"sample"`   | 1     | 1200   | 1100   | 1000   | 5000   | -302   |
| ...            | ...          | ...   | ...    | ...    | ...    | ...    | ...    |
| `"file_1.did"` | `"standard"` | 0     | 1300   | 1100   | 1000   | 5000   | -260   |
| `"file_1.did"` | `"standard"` | 1     | 1200   | 1120   | 1020   | 5020   | -230   |

to the following output:

| `file_id`      | `file_datetime`     | cycle | s44  | s45  | s46  | s47  | s54  | r44  | r45  | r46  | r47  | r54  |
| `"file_1.did"` | 2019-03-01 12:00:00 | 1     | 1200 | 1100 | 1000 | 5000 | -302 | 1250 | 1110 | 1010 | 5010 | -245 |


```{r}
sprd <- bgds %>%
  spread_match(method = "normal") # "linterp")
```

### calculate little delta's and clumped values

Now we can go to the bread and butter of clumpedr, delta calculations!

```{r}
dlts <- sprd %>%
  delta_values(plot_info = didinfo, genplot = TRUE)
```

### collapse cycles: calculate averages and standard deviations

```{r}
coll <- dlts %>%
  collapse_cycles()
```

### append metadata

This just `left_join`s the metadata based on `file_id`, so that we can use it
for outlier removal etc.

```{r}
dati <- coll %>%
  add_info(didinfo)
```

### remove outliers

Based on several criteria, we can get rid of outliers. This needs to happen
before the Empirical Reference Frame is calculated and applied.

```{r}
rout <- dati %>%
  remove_outliers(genplot = TRUE)
```

### empirical transfer function

```{r}
detf <- rout %>%
  # TODO: this currently throws an error due to something in append_expected_valuse()
  # Error: Column `exp` must be length 5180 (the number of rows) or one, not 0
  empirical_transfer_function(session = Preparation, genplot = TRUE)
```

### acid fractionation factor
```{r}
daff <- detf %>%
  acid_fractionation()
```

### temperature calculation
```{r}
temp <- daff %>%
  temperature_calculation() %>%
  pipe_plot(plot_delta)
```

## plots and summary

We can create some summary plots, i.e. of the temperature per standard:

```{r}
# create a tibble that holds heights for text annotations
summ <- temp %>%
  group_by(broadid) %>%
  summarise(y = max(D47_final, na.rm = TRUE) + .05, n = n())

temp %>%
  plot_base(x = broadid, y = D47_final) +
  geom_violin(aes(group = broadid, fill = broadid), alpha = 0.2) +
  geom_jitter(width = .05, alpha = .6) +
  geom_text(aes(x = broadid, y = y, label = n), data = summ, inherit.aes = FALSE)
```

If you don't understand any of the steps, look at the function documentation
(e.g.: `?empirical_transfer_function`) or look at the source code (type the
function name without parentheses into the command line).

Enjoy!
